\begin{center}
  \textsc{Acknowledgement}
\end{center}
\noindent

The end of my studies marks the close of another chapter in my life. Producing and correcting the results presented in this thesis required sustained effort from me, and it also demanded time, patience, and support from those around me. It was, nonetheless, a wonderful period, and I would like to thank everyone who contributed to it.

This thesis lies at the intersection of algebraic topology and the theory of operator algebras. I would therefore like to begin by thanking my lecturers Cathérine Meusburger, Gandalf Lechner, Peter Fiebig, Thomas Creutzig, and Hermann Schulz-Baldes for their excellent courses on the topics essential to this work.

The department also made it possible for me to finance my mathematics studies by offering numerous tutoring positions in linear algebra, analysis, topology, functional analysis, and engineering mathematics. I am grateful to the professors and lecturers Karl-Hermann Neeb, Cathérine Meusburger, Kang Li, and Michael Fried for giving me these opportunities.

I would like to offer my special thanks to Aidan Sims, who delivered a series of lectures on the homology of ample groupoids at our university while I was working on this thesis. Through discussions during the lectures, as well as our evening conversations over dinner, I gained substantial insight into the subject and learned a great deal from him. His stay at our university was both highly informative and genuinely enjoyable for me. Thank you very much.

I would also like to express my gratitude towards Christian Bönicke. Christian took the time to discuss my results once they were written down, suggested non-trivial examples, and helped me improve this thesis through careful listening and well-timed, practical advice. The evening at Alter Simpl and our conversations were both instructive and enjoyable. Thank you very much.

I would like to thank my two supervisors, Kang Li and my mentor in the Master's programme, Cathérine Meusburger, as well as once again Karl-Hermann Neeb. I could always approach them with questions, and they would respond quickly. They also provided valuable guidance on my next career steps. It was a pleasure working with you, and I truly enjoyed our time together. Thank you very much.

I would like to thank the members of the teaching staff Christian van de Ven, Ricardo Correa da Silva, Michael Preeg, Tobias Simon, and Matthias Bauer for a wonderful time.

Finally, I would like to express my sincere gratitude to my partner, Luciana Diaconescu, who had to listen to all of this. She will soon be writing her thesis, and I sincerely hope that I can repay her with the same patience she has shown me and my moods.